Dynamics of thin liquid film flows studied with the lattice Boltzmann method

### 2024-01-16 chapters/Paper_one.tex
Updates from Overleaf

\textit{\small{This chapter has been published and can be found as: Zitz, S., Scagliarini, A., Maddu, S., Darhuber, A. A., \& Harting, J. (2019). Lattice Boltzmann method for thin-liquid-film hydrodynamics. Physical Review E, 100(3), 033313}}
\\
\\

We propose a novel approach to the numerical simulation of thin film flows, based on the lattice Boltzmann method. We outline the basic features of the method, show in which limits the expected thin film equations are recovered and perform validation tests.
The numerical scheme is applied to the viscous Rayleigh-Taylor instability of a thin film and to the spreading of a sessile drop towards its equilibrium contact angle configuration. 
We show that the Cox-Voinov law is satisfied, and that the effect of a tunable slip length on the substrate is correctly captured. 
We address, then, the problem of a droplet sliding on an inclined plane, finding that the Capillary number scales linearly with the Bond number, in agreement with experimental results. 
At last, we demonstrate the ability of the method to handle heterogenous and complex systems by showcasing the controlled dewetting of a thin film on a chemically structured substrate.

\section{Introduction}
Thin layers of liquids on solid surfaces are frequently encountered in a host of natural and technological settings \cite{degennesCapillarityWettingPhenomena2004,fockeLabonaFoilMicrofluidicsThin2010}. 
Therefore, understanding and controlling their stability and dynamics is a central problem for fundamental physics, as well as for applied research in process engineering and nanotechnology~\cite{oronLongscaleEvolutionThin1997,utadaDrippingJettingDrops2007}. 
Coating processes, for instance, rely crucially on the mutual affinity of liquid and surface (i.e. on wettability properties). When the liquid film is sufficiently thin, in fact, it can become unstable, leading to the dewetting of the coated area~\cite{bonnWettingSpreading2009}. 
From the modelling point of view, the challenge consists in the fact that the physics of thin films is intrinsically multiscale, for it involves phenomena ranging from the molecular scale at the three phase contact line, to the micro-/nano-metric size of the film thickness, to the size of the film as a whole, extending over the coated substrate area.

A fully resolved bottom-up atomistic approach would be, obviously, unfeasible, if hydrodynamic regimes are to be explored. 
It clearly appears that some degree of model order reduction is required. 
Most hydrodynamic models of thin liquid films, in the framework of the lubrication theory, simplify the complexity of the full 3D Navier-Stokes equations~\cite{claude-louis-marie-henriMemoireLoisMouvement1827, stokesSteadyMotionIncompressible1848} to one scalar transport equation (the lubrication equation) for the film thickness field $h(x,t)$~\cite{reynoldsTheoryLubricationIts1886,oronLongscaleEvolutionThin1997,crasterDynamicsStabilityThin2009,mitlinDewettingSolidSurface1993}:
\begin{equation}\label{eq:lubr}
    \partial_t h = \nabla \cdot \left(Q(h)\nabla p_{\mbox{\tiny{film}}}\right)
\end{equation}
Here, $Q(h)$ is the mobility function, whose explicit form depends on the boundary condition for the velocity at the surface (for a no-slip boundary, $Q(h) = h^3/(3 \mu)$, with $\mu$ being the dynamic viscosity), and $p_{\mbox{\tiny{film}}}$ is the film pressure at the free liquid surface.
Stable and reliable direct numerical simulations of Eq.~(\ref{eq:lubr}) require sophisticated numerical methods, whose execution is often computationally expensive~\cite{beckerComplexDewettingScenarios2003}.
Moreover, an ever-growing number of microfluidic problems requires to cope with complex fluids rather than simple liquids, i.e. fluids with non-trivial internal microstructure and/or complex non-Newtonian rheological behaviour (e.g. colloidal suspensions, polymer solutions, etc.). 
The quest for an efficient multiscale numerical method for simulating thin film hydrodynamics, versatile for the inclusion of multiphysics features, is, thus, an ongoing endeavour.

In this paper, we present a novel approach to the numerical study of thin liquid films, based on the lattice Boltzmann method (LBM) \cite{succiLatticeBoltzmannEquation2001}. 
Due to the built-in properties of the LBM, our method enjoys an outstanding computational performance, especially on parallel architectures and graphics processing units.

The paper is organized as follows. 
We first present the numerical model and discuss the equations of motion for the hydrodynamic fields that the model covers. 
We then show that these equations effectively correspond, under certain limits, to the lubrication equation of Reynolds.
In section~\ref{sec:results} we present validation results including the Rayleigh-Taylor instability of thin fluid films, the spreading of a sessile droplet on a substrate and the sliding of a droplet on an inclined plane. 
After showcasing the ability of our method to handle large and heterogeneous substrates, we present some computational aspects including the performance of our implementation for Graphics Processing Units (GPUs). 
An appendix is added to provide numerical tests of the validity of the correspondence with lubrication theory (appendix~\ref{app:only}).

@@

When a layer of fluid is characterized by a vertical length scale $H$ much smaller than the longitudinal one $L$, the equations of motion can be simplified under the approximation that the ratio of the length scales, $\varepsilon \equiv H/L$, is small ($\varepsilon \ll1$, see Fig.~\ref{fig:scheme}).
\begin{figure}
    \centering
    \includegraphics[width = 0.75\textwidth]{graphics/Fig_1_Scheme_V3_filters_2.png}
    \caption{Schematic sketch of a model system: a thin liquid film deposited on a flat substrate. The air-liquid interface is represented by the \textit{height} $h(x,y,t)$. The characteristic thickness of the film is given by $H$.}
    \label{fig:scheme}
\end{figure}
In this limit, and for small reduced Reynolds number, $\varepsilon^2Re$ the lubrication approximation tells that the dynamics is governed by equation (\ref{eq:lubr}).
Instead of directly solving Eq.~(\ref{eq:lubr}) numerically, we follow an alternative strategy.
We build our numerical model on a class of LBMs originally proposed as solvers for the shallow water equations~\cite{salmonLatticeBoltzmannMethod1999,dellarNonhydrodynamicModesPriori2002,zhouLatticeBoltzmannMethods2004,vanthangStudy1DLattice2010}.
The lattice Boltzmann equation for the discrete probability density functions of a fluid system subject to a total force (that can include both internal and external forces) $\mathbf{F}_{\mbox{\tiny{tot}}}$, $f_l(\mathbf{x},t)$, reads:
\begin{equation}\label{eq:LBE}
\begin{split}
&f_l(\mathbf{x}+\mathbf{c}^{(l)}\Delta t,t+\Delta t) = \\
&(1 - \omega) f_l(\mathbf{x},t) + \omega f_l^{(eq)}(\mathbf{x},t) + w_l \frac{\Delta t}{c_s^2} \mathbf{c}^{(l)} \cdot \mathbf{F}_{\mbox{\tiny{tot}}},
\end{split}
\end{equation}
where $l$ labels the lattice velocities $\mathbf{c}_l$ and runs from $0$ to $Q-1$, with $Q$ being the number of velocities characterizing the scheme.
Algorithmically, this equation can be seen as made up of two steps. 
A \textit{local collision} step where the $f_l(\mathbf{x},t)$ ``relax'' towards the local equilibrium distributions $f^{(eq)}_l(\mathbf{x},t)$ with rate $\omega = \Delta t/\tau$ (where $\tau$, the relaxation time, is proportional to the kinematic viscosity $\nu$): the distribution functions are substituted by their weighted average (with weights $\omega$ and $1-\omega$) with the equilibria, with an added so-called "source" term (the last term on the right hand side of Eq. (\ref{eq:LBE})), when a force is present. 
A \textit{non-local streaming} step where the updated distribution functions are scattered to the nearest neighbouring sites. 
The parameters $c_s$ (the lattice speed of sound) and $w_l$ (the so called "weights") depend on the geometry of the lattice and are determined under suitable constraints on the form of the tensorial moments in the lattice velocities up to fourth order~\cite{wolf-gladrowLatticeGasCellularAutomata2004}.
We work with two-dimensional square lattices of side length $N\Delta x$, with lattice constant $\Delta x$ and $Q=9$.
For simplicity, we keep $\Delta t=\Delta x=1$ throughout this paper and follow the standard notation, where $c_s = 1/\sqrt{3}$ and the $\mathbf{c}^{(l)} = (c^{(l)}_x,c^{(l)}_y)$, $l=0,1,\dots,8$, are~\cite{qianFractionalPropagationElimination1997,shanLatticeBoltzmannModel1993}

@@

where the left hand side coincides with the momentum flux of the shallow water equation, with the term $gh^2/2$ being the hydrostatic pressure in a thin fluid layer at rest~\cite{dellarNonhydrodynamicModesPriori2002}.

@@

The multiscale Chapman-Enskog expansion \cite{chapmanMathematicalTheoryNonuniform1990,enskogKinetischeTheorieVorgange1917} of such a LBM yields (for small ratios $Ma/Fr$ of the Mach, $Ma=u/c_s$, and Froude, $Fr = u/\sqrt{gH}$, numbers, corresponding also to $\sqrt{g H}/c_s \ll 1$) the following equations for the height and velocity fields~\cite{dellarNonhydrodynamicModesPriori2002,vanthangStudy1DLattice2010,salmonLatticeBoltzmannMethod1999}

@@

where $\nu$, the kinematic viscosity, is related to the relaxation rate $\omega$ appearing in (\ref{eq:LBE}) via $\nu = c_s^2((2-\omega)/2\omega)\Delta t$. 
For stability of the scheme, the condition $Fr < 1$ is also required, which is fulfilled in all our applications, given the low values of $u$ (as discussed in more detail later on).

@@

p_{\mbox{\tiny{film}}} = - \gamma (\nabla^2 h -\Pi(h))

@@

and $\rho_0$ is the (constant) liquid density (equal to $1$, in LBM units). 

@@

These are commonly chosen values in the literature~\cite{moultonEffectDisjoiningPressure2013,oronLongscaleEvolutionThin1997} that correspond to a standard $6-12$ Lennard-Jones intermolecular potential~\cite{fischerExistencePositiveSolutions2018}, though other pairs $(n,m)$ can be used (e.g.~$(2,3), (3,6), (4,10)$~\cite{diezStabilityFinitelengthRivulet2009,wilczekSlidingDropsEnsemble2017, wedershovenInfraredLaserInduced2014}). 

@@

The film pressure is specific to model thin film dynamics, in general however one can make use of other force terms e.g. to couple fluid layers which has been shown in \cite{roccaDevelopmentLatticeBoltzmann2012}. 
The second term on the right hand side in Eq.~(\ref{eq:force}) introduces a friction with the substrate of the form

@@

Here, $\delta$ acts as a regularizing parameter, which can be identified with an effective slip length. 
Finally, the last term in Eq.~(\ref{eq:force}), $\mathbf{F}$ accounts for any other possible source of forcing (e.g. the gravity component parallel to the substrate in the case of a liquid film deposited on an inclined plate).
Equipped with such extra terms, equations (\ref{eq:hydro}) become

@@

Let us notice at this point that for most microfluidic applications we are actually interested in, the advection term on the left hand side of the second equation of (\ref{eq:hydro2}) is indeed negligible as compared to the right hand side (the Reynolds number \cite{reynoldsIIIExperimentalInvestigation1883,sommerfeldBeitrag1909} being much smaller than one).  
Analogously, the longitudinal viscous terms $\nu \nabla^2 (h\mathbf{u})$ and $2\nu \nabla (\nabla \cdot (h\mathbf{u}))$ are of order $\varepsilon^2$ smaller in the ratio of length scales than the friction term $\nu \alpha(h) \mathbf{u}$ (since the former scale as $\nu H\frac{U}{L^2}$, whereas the latter as $\nu \frac{U}{H}$).

@@

The validity of these considerations has been numerically tested in some selected cases (representative of typical applications) and the results are shown and discussed in the appendix.

@@

For processes evolving on time-scales $\tilde{t}$ such that $\tilde{t}\gg \frac{h}{\alpha(h)\nu}$, one can consider the ``quasi-steady'' limit of the second of these equations (setting $\partial_t (h\mathbf{u}) \approx 0$), which yields

@@

In the no slip limit, $\delta \rightarrow 0$, and in absence of gravity and other forces, Eq.~(\ref{eq:uslaved}) simplifies into

@@

In essence, our method is, therefore, an alternative solver of the lubrication equation (at least in the inertialess regime, $Re \ll 1$, and for very thin films, $\varepsilon \ll 1$), that brings in, from the computational point of view, the added values of excellent scalability of the corresponding LBM algorithm on parallel architectures, as we shall see in the following sections. 
Similar ideas have also been developed for reaction-diffusion equations~\cite{caliDiffusionHydrodynamicDispersion1992,kingdonReactionflowLatticeBoltzmann1992,chenLatticeMethodsTheir1995,weimarNonlinearReactionsAdvected1996} and the modelling of surface tension effects by gradients of auxiliary fields~\cite{swiftLatticeBoltzmannSimulations1996, lishchukLatticeBoltzmannAlgorithm2003} based on the color gradient method~\cite{gunstensenLatticeBoltzmannModel1991}.

Before concluding this section, let us notice that special care has to be taken in the implementation of the numerical scheme, when evaluating the forcing term since it contains higher order derivatives (the gradient $p_{\mbox{\tiny{film}}}$, which in turn includes the Laplace pressure $\gamma \nabla^2 h$, see Eq.(~\ref{eq:pfilm}) and, hence, spurious lattice effects may arise.  
We noticed, for example, that a centered scheme to calculate gradients \cite{zhouLatticeBoltzmannMethods2004} does not guarantee the sufficient degree of isotropy on the lattice as, e.g., for the relaxation of a droplet (discussed in section~\ref{sec:results}), where it led to unphysical droplet shapes.
Therefore, we use the following expressions to compute the gradients
\begin{equation}\label{eq:derivative}
    \nabla \phi(\mathbf{x}) = 3\sum_{l=0}^8w_l\mathbf{c}^{(l)} \phi(\mathbf{x}+\mathbf{c}^{(l)})+O(\nabla^3),
\end{equation}
and the Laplacian 
\begin{align}\label{eq:n_laplace}
    \nabla^2 \phi(\mathbf{x}) =& \frac{1}{6}\bigg[4\sum_{l = \text{odd}} w_l \phi(\mathbf{x}+\mathbf{c}^{(l)})\nonumber\\ 
    &+ 1\sum_{l = \text{even}} w_l \phi(\mathbf{x}+\mathbf{c}^{(l)}) - 20\phi(\mathbf{x}) \bigg] + O(\nabla^4),
\end{align}
respectively \cite{junkDiscretizationsIncompressibleNavier2000,thampiIsotropicDiscreteLaplacian2013}, for a generic scalar field $\phi$ (be it the height field $h$, the pressure $p_{\mbox{\tiny{film}}}$ or a position dependent surface tension field).
Besides the higher degree of the isotropy, the scheme (\ref{eq:derivative}-\ref{eq:n_laplace}) has the advantage of employing directly the set of lattice Boltzmann speeds.

@@

\begin{figure}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_2_1_rti_renders_a).png}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_2_2_rti_renders_b).png}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_2_3_rti_renders_c).png}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_2_4_rti_renders_d).png}
  \caption{Time evolution of the free surface for the Rayleigh-Taylor instability at $\tau_{\mbox{\tiny{cap}}}\approx50, 75, 100, 188$, corresponding to the lattice Boltzmann time steps [9000, 14000, 19000, 35000] $\Delta t$ in (a) - (d) respectively. For a more clear visualization we only show a small patch of size $256\times256$ centered in the middle of the $2048\times 2048$ domain. The fluctuations of earlier states still follow the linear stability analysis (See Fig.~\ref{fig:RTI} for the the power spectrum  of the  height fluctuations versus wavenumber).}
  \label{fig:RTI_evolution}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{graphics/Fig_3_new_RTI_spectra_single_color_same_tau_capkc_correct_rescaled_x_axis_inset.png}
    \caption{Power spectrum of the height fluctuations versus wavenumber.} The different colors and symbols belong to different values of graviational acceleration, $g=4\cdot 10^{-5}$ is given by blue circles (\textcolor{pyblue}{$\bullet$}), $g=6\cdot 10^{-5}$ by orange triangles (\textcolor{pyorange}{$\blacktriangle$}) and $g=8\cdot 10^{-5}$ is given by green squares (\textcolor{pygreen}{$\blacksquare$}). Same colored lines are taken at different time steps. In the inset we show the growth rate $\sigma(k)$ for the largest value of $g$ (symbols) and the theoretical growth rate according to Eq.~(\ref{eq:RTgrowth}) (solid line).
    \label{fig:RTI}
\end{figure}

>>>>

\textit{\small{This chapter has been published as: Zitz, S., Scagliarini, A., Maddu, S., Darhuber, A. A., \& Harting, J. (2019). Lattice Boltzmann method for thin-liquid-film hydrodynamics. Physical Review E, 100(3), 033313}}
\\
\\

We propose an approach to the numerical simulation of thin film flows based on the lattice Boltzmann method.
We outline the basic features of the method, show in which limits the expected thin-film equations are recovered, and perform validation tests.
The numerical scheme is applied to the viscous Rayleigh-Taylor instability of a thin film and to the spreading of a sessile drop towards its equilibrium contact angle configuration. 
We show that the Cox-Voinov law is satisfied, and that the effect of a tunable slip length on the substrate is correctly captured. 
We address, then, the problem of a droplet sliding on an inclined plane, finding that the Capillary number scales linearly with the Bond number, in agreement with experimental results. 
At last, we demonstrate the ability of the method to handle heterogenous and complex systems by showcasing the controlled dewetting of a thin film on a chemically structured substrate.

\section{Introduction}
Thin layers of liquids on solid surfaces are frequently encountered in a host of natural and technological settings \cite{degennesCapillarityWettingPhenomena2004,fockeLabonaFoilMicrofluidicsThin2010}. 
Therefore, understanding and controlling their stability and dynamics is a central problem for fundamental physics, as well as for applied research in process engineering and nanotechnology~\cite{oronLongscaleEvolutionThin1997,utadaDrippingJettingDrops2007}. 
Coating processes, for instance, rely crucially on the mutual affinity of liquid and surface (i.e. on wettability properties). 
When the liquid film is sufficiently thin, in fact, it can become unstable, leading to the dewetting of the coated area~\cite{bonnWettingSpreading2009}. 
From the modelling point of view, the challenge consists in the fact that the physics of thin films is intrinsically multiscale, for it involves phenomena ranging from the molecular scale at the three-phase contact line, to the micro- and nanometric size of the film thickness to the size of the film as a whole, extending over the coated substrate area.

A fully resolved bottom-up atomistic approach would be, obviously, unfeasible if hydrodynamic regimes are to be explored. 
It clearly appears that some degree of model order reduction is required. 
Most hydrodynamic models of thin liquid films, in the framework of the lubrication theory, simplify the complexity of the full three-dimensional (3D) Navier-Stokes equations~\cite{claude-louis-marie-henriMemoireLoisMouvement1827, stokesSteadyMotionIncompressible1848} to one scalar transport equation (the lubrication equation) for the film thickness field $h(\mathbf{x},t)$~\cite{reynoldsTheoryLubricationIts1886,oronLongscaleEvolutionThin1997,crasterDynamicsStabilityThin2009,mitlinDewettingSolidSurface1993}:
\begin{equation}\label{eq:lubr}
    \partial_t h = \nabla \cdot \left[Q(h)\nabla p_{\mbox{\tiny{film}}}\right]
\end{equation}
Here, $Q(h)$ is the mobility function, whose explicit form depends on the boundary condition for the velocity at the surface [for a no-slip boundary, $Q(h) = h^3/(3 \mu)$, with $\mu$ being the dynamic viscosity], and $p_{\mbox{\tiny{film}}}$ is the film pressure at the free liquid surface.
Stable and reliable direct numerical simulations of Eq.~(\ref{eq:lubr}) require sophisticated numerical methods, whose execution is often computationally expensive~\cite{beckerComplexDewettingScenarios2003}.
Moreover, an ever-growing number of microfluidic problems requires us to cope with complex fluids rather than simple liquids, i.e., fluids with nontrivial internal microstructure and/or complex non-Newtonian rheological behaviour (e.g., colloidal suspensions, polymer solutions, etc.). 
The quest for an efficient multiscale numerical method for simulating thin-film hydrodynamics, versatile for the inclusion of multiphysics features, is, thus, an ongoing endeavor.

In this paper, we present an approach to the numerical study of thin liquid films based on the lattice Boltzmann method (LBM)~\cite{succiLatticeBoltzmannEquation2001}. 
Due to the built-in properties of the LBM, our method enjoys an outstanding computational performance, especially on parallel architectures and graphics processing units (GPUs).

The paper is organized as follows. 
We first present the numerical model and discuss the equations of motion for the hydrodynamic fields that the model covers. 
We then show that these equations effectively correspond, under certain limits, to the lubrication equation of Reynolds.
In Sec.~\ref{sec:results} we present validation results, including the Rayleigh-Taylor instability of thin-fluid-films, the spreading of a sessile droplet on a substrate, and the sliding of a droplet on an inclined plane. 
After showcasing the ability of our method to handle large and heterogeneous substrates, we present some computational aspects including the performance of our implementation for GPUs. 
An appendix is added to provide numerical tests of the validity of the correspondence with lubrication theory (Appendix).

@@

When a layer of fluid is characterized by a vertical length scale $H$ much smaller than the longitudinal one $L$, the equations of motion can be simplified under the approximation that the ratio of the length scales, $\varepsilon \equiv H/L$, is small ($\varepsilon \ll1$, see Fig.~\ref{fig:scheme}).
\begin{figure}
    \centering
    \includegraphics[width = 0.75\textwidth]{graphics/Fig_1_Scheme_V3_filters_2.png}
    \caption{Schematic sketch of a model system: a thin liquid film deposited on a flat substrate. The air-liquid interface is represented by the \textit{height} $h(x,y,t)$. The characteristic thickness of the film is given by $H$.}
    \label{fig:scheme}
\end{figure}
In this limit, and for small reduced Reynolds number, $\varepsilon^2\text{Re}$ (where $\text{Re} = \frac{UL}{\nu}$, with $U$ being a characteristic velocity of the fluid system and $\nu$ being the fluid's kinematic viscosity), the lubrication approximation tells that the dynamics is governed by Eq.~(\ref{eq:lubr}).
Instead of directly solving Eq.~(\ref{eq:lubr}) numerically, we follow an alternative strategy.
We build our numerical model on a class of LBMs originally proposed as solvers for the shallow water equations~\cite{salmonLatticeBoltzmannMethod1999,dellarNonhydrodynamicModesPriori2002,zhouLatticeBoltzmannMethods2004,vanthangStudy1DLattice2010}.
The lattice Boltzmann equation for the discrete probability density functions of a fluid system subject to a total force (that can include both internal and external forces) $\mathbf{F}_{\mbox{\tiny{tot}}}$, $f_l(\mathbf{x},t)$, reads:
\begin{equation}\label{eq:LBE}
\begin{split}
&f_l(\mathbf{x}+\mathbf{c}^{(l)}\Delta t,t+\Delta t) = \\
&(1 - \omega) f_l(\mathbf{x},t) + \omega f_l^{(eq)}(\mathbf{x},t) + w_l \frac{\Delta t}{c_s^2} \mathbf{c}^{(l)} \cdot \mathbf{F}_{\mbox{\tiny{tot}}},
\end{split}
\end{equation}
where $l$ labels the lattice velocities $\mathbf{c}_l$ and runs from $0$ to $Q-1$, with $Q$ being the number of velocities characterizing the scheme.
Algorithmically, this equation can be seen as composed of two steps. 
There is a \textit{local collision} step where the $f_l(\mathbf{x},t)$ ``relax'' towards the local equilibrium distributions $f^{(eq)}_l(\mathbf{x},t)$ with rate $\omega = \Delta t/\tau$ (where $\tau$, the relaxation time, is proportional to the kinematic viscosity $\nu$): The distribution functions are substituted by their weighted average (with weights $\omega$ and $1-\omega$) with the equilibria, with an added "source" term [the last term on the right-hand side of Eq.~(\ref{eq:LBE})], when a force is present. 
There is a \textit{non-local streaming} step where the updated distribution functions are scattered to the nearest-neighbouring sites. 
The parameters $c_s$ (the lattice speed of sound) and $w_l$ (the "weights") depend on the geometry of the lattice and are determined under suitable constraints on the form of the tensorial moments in the lattice velocities up to fourth order~\cite{wolf-gladrowLatticeGasCellularAutomata2004}.
We work with two-dimensional square lattices of side length $N\Delta x$, with lattice constant $\Delta x$ and $Q=9$.
For simplicity, we keep $\Delta t=\Delta x=1$ throughout this paper and follow the standard notation, where $c_s = 1/\sqrt{3}$ and the $\mathbf{c}^{(l)} = [c^{(l)}_x,c^{(l)}_y]$, $l=0,1,\dots,8$, are~\cite{qianFractionalPropagationElimination1997,shanLatticeBoltzmannModel1993}

@@

where the left-hand side coincides with the momentum flux of the shallow-water equation, with the term $gh^2/2$ being the hydrostatic pressure in a thin fluid layer at rest~\cite{dellarNonhydrodynamicModesPriori2002}.

@@

The multiscale Chapman-Enskog expansion \cite{chapmanMathematicalTheoryNonuniform1990,enskogKinetischeTheorieVorgange1917} of such an LBM yields (for small ratios $Ma/Fr$ of the Mach, $Ma=u/c_s$, and Froude, $Fr = u/\sqrt{gH}$, numbers, corresponding also to $\sqrt{g H}/c_s \ll 1$) the following equations for the height and velocity fields~\cite{dellarNonhydrodynamicModesPriori2002,vanthangStudy1DLattice2010,salmonLatticeBoltzmannMethod1999}

@@

where $\nu$, the kinematic viscosity, is related to the relaxation rate $\omega$ appearing in (\ref{eq:LBE}) via $\nu = c_s^2[(2-\omega)/2\omega]\Delta t$. 
For stability of the scheme, the condition $Fr < 1$ is also required, which is fulfilled in all our applications, given the low values of $u$ (as discussed in more detail later).

@@

p_{\mbox{\tiny{film}}} = - \gamma [\nabla^2 h -\Pi(h)]

@@

and $\rho_0$ is the (constant) liquid density (equal to $1$ in LBM units). 

@@

These are commonly chosen values in the literature~\cite{moultonEffectDisjoiningPressure2013,oronLongscaleEvolutionThin1997} that correspond to a standard $6-12$ Lennard-Jones intermolecular potential~\cite{fischerExistencePositiveSolutions2018}, though other pairs $(n,m)$ can be used [e.g.~$(2,3), (3,6), (4,10)$]~\cite{diezStabilityFinitelengthRivulet2009,wilczekSlidingDropsEnsemble2017, wedershovenInfraredLaserInduced2014}. 

@@

The film pressure is specific to model thin-film dynamics; in general, however, one can make use of other force terms, e.g., to couple fluid layers which has been shown in Ref.~\cite{roccaDevelopmentLatticeBoltzmann2012}. 
The second term on the right-hand side in Eq.~(\ref{eq:force}) introduces a friction with the substrate of the form

@@

Here $\delta$ acts as a regularizing parameter, which can be identified with an effective slip length. 
Finally, the last term in Eq.~(\ref{eq:force}), $\mathbf{F}$, accounts for any other possible source of forcing (e.g., the gravity component parallel to the substrate in the case of a liquid film deposited on an inclined plate).
Equipped with such extra terms, Eqs.~(\ref{eq:hydro}) become

@@

Let us notice at this point that for most microfluidic applications we are actually interested in, the advection term on the left-hand side of the second equation of (\ref{eq:hydro2}) is indeed negligible as compared to the right-hand side (the Reynolds number \cite{reynoldsIIIExperimentalInvestigation1883,sommerfeldBeitrag1909} being much smaller than one).  
Analogously, the longitudinal viscous terms $\nu \nabla^2 (h\mathbf{u})$ and $2\nu \nabla [\nabla \cdot (h\mathbf{u})]$ are of order $\varepsilon^2$ smaller in the ratio of length scales than the friction term $\nu \alpha(h) \mathbf{u}$ (since the former scale as $\nu H\frac{U}{L^2}$, whereas the latter as $\nu \frac{U}{H}$).

@@

The validity of these considerations has been numerically tested in some selected cases (representative of typical applications) and the results are shown and discussed in the Appendix.

@@

For processes evolving on timescales $\tilde{t}$ such that $\tilde{t}\gg \frac{h}{\alpha(h)\nu}$, one can consider the ``quasisteady'' limit of the second of these equations [setting $\partial_t (h\mathbf{u}) \approx 0$], which yields

@@

In the no-slip limit, $\delta \rightarrow 0$, and in absence of gravity and other forces, Eq.~(\ref{eq:uslaved}) simplifies into

@@

In essence, our method is, therefore, an alternative solver of the lubrication equation (at least in the inertialess regime, $\text{Re} \ll 1$, and for very thin films, $\varepsilon \ll 1$), that brings in, from the computational point of view, the added values of excellent scalability of the corresponding LBM algorithm on parallel architectures, as we shall see in the following sections. 
Similar ideas have also been developed for reaction-diffusion equations~\cite{caliDiffusionHydrodynamicDispersion1992,kingdonReactionflowLatticeBoltzmann1992,chenLatticeMethodsTheir1995,weimarNonlinearReactionsAdvected1996} and the modeling of surface tension effects by gradients of auxiliary fields~\cite{swiftLatticeBoltzmannSimulations1996, lishchukLatticeBoltzmannAlgorithm2003} based on the color gradient method~\cite{gunstensenLatticeBoltzmannModel1991}.

Before concluding this section, let us notice that special care has to be taken in the implementation of the numerical scheme, when evaluating the forcing term since it contains higher order derivatives [the gradient $p_{\mbox{\tiny{film}}}$, which in turn includes the Laplace pressure $\gamma \nabla^2 h$, see Eq.~(\ref{eq:pfilm})] and, hence, spurious lattice effects may arise.  
We noticed, for example, that a centered scheme to calculate gradients \cite{zhouLatticeBoltzmannMethods2004} does not guarantee the sufficient degree of isotropy on the lattice as, e.g., for the relaxation of a droplet (discussed in Sec.~\ref{sec:results}), where it led to unphysical droplet shapes.
Therefore, we use the following expressions to compute the gradients:
\begin{equation}\label{eq:derivative}
    \nabla \phi(\mathbf{x}) = 3\sum_{l=0}^8w_l\mathbf{c}^{(l)} \phi(\mathbf{x}+\mathbf{c}^{(l)})+O(\nabla^3),
\end{equation}
and the Laplacian 
\begin{align}\label{eq:n_laplace}
    \nabla^2 \phi(\mathbf{x}) =& \frac{1}{6}\bigg[4\sum_{l = \text{odd}} w_l \phi(\mathbf{x}+\mathbf{c}^{(l)})\nonumber\\ 
    &+ 1\sum_{l = \text{even}} w_l \phi(\mathbf{x}+\mathbf{c}^{(l)}) - 20\phi(\mathbf{x}) \bigg] + O(\nabla^4),
\end{align}
respectively~\cite{junkDiscretizationsIncompressibleNavier2000,thampiIsotropicDiscreteLaplacian2013}, for a generic scalar field $\phi$ (be it the height field $h$, the pressure $p_{\mbox{\tiny{film}}}$ or a position-dependent surface tension field).
Besides the higher degree of the isotropy, the scheme (\ref{eq:derivative}) and (\ref{eq:n_laplace}) has the advantage of employing directly the set of lattice Boltzmann speeds.

@@

\begin{figure}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_2_1_rti_renders_a).png}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_2_2_rti_renders_b).png}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_2_3_rti_renders_c).png}
    \includegraphics[width=0.48\textwidth]{graphics/Fig_2_4_rti_renders_d).png}
  \caption{Time evolution of the free surface for the Rayleigh-Taylor instability at $\tau_{\mbox{\tiny{cap}}}\approx 50, 75, 100, 188$, corresponding to the lattice Boltzmann time steps [9000, 14000, 19000, 35000] $\Delta t$ in (a) - (d). For a more clear visualization we only show a small patch of size $256\times256$ centered in the middle of the $2048\times 2048$ domain. The fluctuations of earlier states still follow the linear stability analysis (See Fig.~\ref{fig:RTI} for the the power spectrum of the  height fluctuations versus wave number).}
  \label{fig:RTI_evolution}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{graphics/Fig_3_new_RTI_spectra_single_color_same_tau_capkc_correct_rescaled_x_axis_inset.png}
    \caption{Power spectrum of the height fluctuations versus wave number. The different colors and symbols belong to different values of graviational acceleration; $g=4\cdot 10^{-5}$ is given by blue circles (\textcolor{pyblue}{$\bullet$}), $g=6\cdot 10^{-5}$ by orange triangles (\textcolor{pyorange}{$\blacktriangle$}) and $g=8\cdot 10^{-5}$ is given by green squares (\textcolor{pygreen}{$\blacksquare$}). Same colored lines are taken at different time steps. In the inset we show the growth rate $\sigma(k)$ for the largest value of $g$ (symbols) and the theoretical growth rate according to Eq.~(\ref{eq:RTgrowth}) (solid line).}
    \label{fig:RTI}
\end{figure}

The Rayleigh-Taylor instability occurs when a denser fluid is accelerated against a less dense one~\cite{rayleighInvestigationCharacterEquilibrium1882,taylorInstabilityLiquidSurfaces1997,kullTheoryRayleighTaylorInstability1991,sharpOverviewRayleighTaylorInstability1984}. 
This can be the case, for instance, for a liquid film coating a ceiling, under the action of gravity. 
In such a configuration gravity tends, of course, to deform (and eventually disrupt) the film, while surface tension has a stabilizing effect.
As a result of these competing mechanisms, any surface perturbation is stable or unstable depending on whether its characteristic wave number $k$ is smaller or larger than a certain critical value $k_c$. 
Linear stability analysis calculations in the framework of lubrication theory provide the following growth rate $\sigma(k)$:
\begin{equation}\label{eq:RTgrowth}
    \sigma(k) = \frac{\rho g h_0^3}{3\mu}(k^2 - l_{\mbox{\tiny{cap}}}^2 k^4),
\end{equation}
where $l_{\mbox{\tiny{cap}}} = (\gamma/g)^{1/2}$ is the capillary length. 
Unstable (stable) modes correspond to $\sigma(k)>0$ [$\sigma(k)<0$] and the critical wave number is, therefore, such that $\sigma(k_c)=0$, i.e., $k_c = 1/l_{\mbox{\tiny{cap}}}$.
On a lattice of size $2048 \times 2048$ nodes, we initialize the film height according to
\begin{equation}
  h(\mathbf{x},0) = h_0(1 + \varepsilon(\mathbf{x})),
\end{equation}
with $\varepsilon$ a random variable homogeneously distributed in $[1\cdot 10^{-4},-1\cdot 10^{-4}]$ and $h_0 = 1$. 
Forcing should always be below a certain threshold. Thus, for the gravitational acceleration we choose values within the interval $|g| = [4,8]\cdot 10^{-5}$. 
Furthermore, we fix the surface tension to be $\gamma=0.01$. 
This results in critical wave numbers ranging from $k_c= 0.06$ to $k_c = 0.09$.
Figure~\ref{fig:RTI_evolution} shows snapshots of the free surface from various time steps, where the growth of the perturbations is shown as time increases. 
The last panel is already beyond the linear regime.
